---
title: Gun Violence in the US. Application of Unsupervised Learning Methods for Trend Exploration
author:
  - name: Sumaira Afzal
    affiliation: York University School of Continuing Studies
  - name: Viraja Ketkar
    affiliation: York University School of Continuing Studies
  - name: Murlidhar Loka
    affiliation: York University School of Continuing Studies
  - name: Vadim Spirkov
    affiliation: York University School of Continuing Studies
abstract: >
  To do

output:
  rticles::rjournal_article:
    includes:
      in_header: preamble.tex
   
---

```{r echo=FALSE, message=FALSE, warnings=FALSE}
# Clean all variables that might be left by other script to avoid collusion
rm(list=ls(all=TRUE))
# load required libraries
library(ggplot2) # plotting lib
library(dplyr)  # data manipuation
library(RColorBrewer) # color palettes
library(xtable) # tabular data formatting 

# set xtable properties for the project
options(xtable.floating = T)
options(xtable.timestamp = "")
options(xtable.comment = F)

# pick palettes
mainPalette = brewer.pal(8,"Dark2")
```

```{r global_options, include=FALSE}
# make the images flow nicely
knitr::opts_chunk$set(fig.pos = 'H')
```


## Background


## Objective

The objective of this research is to ...


# Data Analysis

The data set used for this research contains 260k of gun violence incidents in the US between January 2013 and March 2018. The data has been soursed from [Kaggle](https://www.kaggle.com/jameslko/gun-violence-data). 

Originallly the data set was uploaded to Kaggle from Gun Violence Archive (GVA)  Web site [ gunviolencearchive.org](https://www.gunviolencearchive.org/). This is a not for profit corporation formed in 2013 to provide free online public access to accurate information about gun-related violence in the United States. GVA will collect and check for accuracy, comprehensive information about gun-related violence in the U.S. and then post and disseminate it online. 


## Data Dictionary


Column Name                 | Column Description  
----------------------------| ------------------- 
incident_id                 | Incident ID
date                        | Date of crime
state                       | State
city_or_countyCity          | City/county of crime
address                     | Address of the location of the crime
n_killed                    | Number of people killed
n_injured                   | Number of people injured
incident_url                | URL regarding the incident
source_url                  | Reference to the reporting source
incident_url_fields_missing | TRUE if the incident_url is present, FALSE otherwise
congressional_district      | Congressional district id
gun_stolen                  | Status of guns involved in the crime (i.e. Unknown, Stolen, etc...)
gun_type                    | Typification of guns used in the crime
incident_characteristics    | Characteristics of the incidence
latitude                    | Location of the incident
location_description        | Description of the location
longitude                   | Location of the incident
n_guns_involved             | Number of guns involved in incident
notes                       | Additional information of the crime
participant_age             | Age of participant(s) at the time of crime (victicms nad suspects)
participant_age_group       | Age group of participant(s) at the time crime
participant_gender          | Gender of participant(s)
participant_name            | Name of participant(s) involved in crime
participant_relationship    | Relationship of participant to other participant(s)
participant_status          | Extent of harm done to the participant
participant_type            | Type of participant (victim or suspect)
sources                     | Participants source
state_house_district        | Voting house district
state_senate_district       | Territorial district from which a senator to a state legislature is elected.


## Data Exploration

```{r message=FALSE, warning=FALSE}

```
  

### Missing Data



#### Takeaways from Data Exploration Excersize

## Data Preparation


### Data Imputing



# Modeling and Evalutation


## Feature Selection


### Data Upsampling



## Partitioning Clustering Approach



## Hierarchical Clustering Approach




## Density-based Clustering Methods



## Clustering Method Evaluation



# Model Deployment


# Conclusion



\bibliography{RJreferences}


# Note from the Authors

This file was generated using [_The R Journal_ style article template](https://github.com/rstudio/rticles), additional information on how to prepare articles for submission is here - [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf). The article itself is an executable R Markdown file that could be [downloaded from Github](https://github.com/ivbsoftware/big-data-final-2/blob/master/docs/R_Journal/big-data-final-2/) with all the necessary artifacts.
